---
title: "Visualise results from frequency estimation and compare with propotions barcode data"
output: pdf_document
---

This script plots results from running COIL extension on TM border data partioned by site and site and year, compared to proportion frequencies, which were based on data excluding all multiclonal. MCMC fit to each partition using, 10000 iterations, $\psi = 1$, $\phi = 1$,$\alpha = 1$, $\beta = 1$, $\sigma^2 = 1$. In the final draft on the manuscript the frequencies inferred under the COIL extension model were deemed superfluous. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
```

```{r, cach = FALSE, echo = FALSE} 
rm(list = ls())
library("arm") # for logit

# Load inferred values (based on data with multiclonal but not missing) 
load('../../../RData/COIext_MOIpop_extracted_site_year.RData')
MOI_quants_site_year <- MOI_quants
load('../../../RData/COIext_f_extracted_site_year.RData')
frequency_quants_site_year <- frequency_quants
load('../../../RData/COIext_MOIpop_extracted_site.RData')
MOI_quants_site <- MOI_quants
load('../../../RData/COIext_f_extracted_site.RData')
frequency_quants_site <- frequency_quants
load('../../../RData/COIext_MOIpop_extracted_all.RData')
MOI_quants_all <- MOI_quants
load('../../../RData/COIext_f_extracted_all.RData')
frequency_quants_all <- frequency_quants

# Based on data with no multiclonal
load('../../../RData/Barcode_frequencies_no_multiclonal.RData')
load('../../../RData/Data_store_Barcode.RData')
nsamples_no_missing <- nrow(Data_store$Data_no_missing)
nsamples_no_missing_mixed <- nrow(Data_store$Data_no_multiclonal_missing)
```

***************************************************

# MOI with site and year

```{r, cach = FALSE, echo = FALSE, include = TRUE} 
site_years <- colnames(MOI_quants_site_year)
sites <- unique(sapply(strsplit(site_years, split = ':'), FUN = function(x){x[1]}))
cols <- rainbow(length(sites)); names(cols) <- sites

# Plot average MOIs
plot(NULL, ylim = c(1,2.5), xlim = c(1,10),
     bty = 'n', xlab = 'Year', ylab = 'Population-level MOI', 
     las = 2, xaxt = 'n')
axis(side = 1, at = 1:10, labels = 2001:2010, las = 2)
for(site in sites){
  ind <- grep(site, colnames(MOI_quants_site_year))
  years <- as.numeric(sapply(strsplit(colnames(MOI_quants_site_year)[ind], split = ':'), FUN = function(x){x[2]}))
  points(y = MOI_quants_site_year['50%', ind], x = years, col = cols[site], pch = 16)
  segments(y0 = MOI_quants_site_year['2.5%', ind], y1 = MOI_quants_site_year['97.5%', ind], 
           x0 = years, x1 = years, 
           col = cols[site])
}
legend('top', legend = c("MKT", "MLA", "MKK", "WPA"), col = cols[sites], 
       pch = 16, bty = 'n', y.intersp = 1.2)

# Is the trend in median posterior estimates significant? Yes. 
X0 <- colnames(MOI_quants_site_year)
X_years <- as.numeric(sapply(strsplit(X0, split = ':'), FUN = function(x){x[2]}))
X_sites <- sapply(strsplit(X0, split = ':'), FUN = function(x){x[1]})
fit_moi_year <- lm(MOI_quants_site_year['50%',] ~ X_years)
summary(fit_moi_year)

# Summary of fitting procedure
# annova(year + sites + interaction) -> interaction not signif
# annova(year + sites) -> sites signif (p-value 0.02); but no site signif in summary(year + sites)
# annova(year) -> years highly signif
fit_moi_sites_year <- lm(MOI_quants_site_year['50%',] ~ X_years + X_sites)
summary(fit_moi_sites_year) # Trend is not significant over sites 
```

```{r, cach = FALSE, echo = FALSE, include = TRUE}
# Not that MOI_quants_site['50%',] has been translated by one
# in order to start y axis at 1
X <- barplot(MOI_quants_site['50%',]-1, col = 'grey', ylim = c(0,1), ylab = 'Population-level MOI', yaxt = 'n', names.arg = c("MKT", "MLA", "MKK", "WPA")) # Checked site order
segments(y0 = MOI_quants_site['2.5%',]-1, y1 = MOI_quants_site['97.5%',]-1,
         x0 = X, x1 = X)
axis(side = 2, at = seq(0,1,0.2), labels = seq(1,2,0.2))
```

# MOI over all 
Average MOI: `r MOI_quants_all[1]`, which is pretty similar to averaging over the site-wise mediums: `r sum(MOI_quants_site['50%',])/4` 


***********************************

# Inferred frequencies by site and year

```{r, cach = FALSE, echo = FALSE, include = TRUE} 

# Plot frequencies against year 
plot(NULL, ylim = c(0,1), xlim = c(1,10), xaxt = 'n',
     bty = 'n', xlab = 'Year', ylab = 'Frequency')
axis(side = 1, at = 1:10, labels = 2001:2010, las = 1)

for(site in sites){
  ind <- grep(site, site_years)
  years_char <- sapply(strsplit(site_years[ind], split = ':'), FUN = function(x){x[2]})
  years <- as.numeric(years_char)
  
  for(i in 1:length(ind)){
    X0 <- rnorm(length(frequency_quants_site_year[,'50%',ind[i]]), sd = 0.2)
    points(y = frequency_quants_site_year[,'50%',ind[i]], 
           x = rep(years[i], 93) + X0, 
           col = adjustcolor(cols[site], alpha.f = 1), pch = 16, cex = 0.5)
    segments(y0 = frequency_quants_site_year[,'2.5%',ind[i]], 
             y1 = frequency_quants_site_year[,'97.5%',ind[i]], 
             x0 = rep(years[i], 93) + X0, 
             x1 = rep(years[i], 93) + X0, 
             col = adjustcolor(cols[site], alpha.f = 0.5), lwd = 0.5)
    points(y = FreqResultsStore_counting$site_years[years_char[i],site,], pch = 4, 
           x = rep(years[i], 93) + X0, cex = 0.5, col = adjustcolor(cols[site], alpha.f = 1))
  }
}
legend('bottom', legend = sites, col = cols[sites], 
       pch = 16, bty = 'n', cex = 0.8)




# Do frequencies vary between sites and years? --------------------------------------
Data_short_format <- frequency_quants_site_year[,'50%',]
sites_short_format <- sapply(strsplit(colnames(Data_short_format), ':'), FUN = function(x){x[1]})
years_short_format <- as.numeric(sapply(strsplit(colnames(Data_short_format), ':'), FUN = function(x){x[2]}))
Data_long_format <- data.frame(freq = as.vector(Data_short_format), 
                               snp_id = as.factor(rep(1:93,ncol(Data_short_format))),
                               site = as.factor(rep(sites_short_format, each = 93)),
                               year = rep(years_short_format, each = 93))

# Fit pooled No
fit_pooled <- lm(freq ~ site + year, data = Data_long_format) 
summary(fit_pooled) # Pooled: site/year not signif 
anova(fit_pooled) # Pooled: site/year not signif 

# Fit individually
snp_ids <- unique(Data_long_format$snp_id)
significant_year <- c()
significant_site <- c()
for(snp_id in snp_ids){
  ind <- Data_long_format$snp_id == snp_id
  Data_subset <- Data_long_format[ind,]
  fit_individual <- lm(freq ~ year + site, data = Data_subset)
  Z <- summary(fit_individual)
  if(Z$coefficients['year','Pr(>|t|)'] < (0.05)/93){ # toggle between 'site' and 'year'
    print(Z$coefficients['year',])
    significant_year <- c(significant_year, snp_id)
  }
}
significant_year
# Differences over time after bonferroni correction: 
```


## Do proportion frequencies vary between sites and years?
This section features in the supplementary of the final draft of the manuscript

```{r, cach = FALSE, echo = FALSE} 
# Based on point estimates of frequencies, since these are what are used in the IBD analyses
rm(Data_long_format)
Freq <- as.vector(FreqResultsStore_counting$site_years)
ind_na <- is.na(Freq)
sites_short_format <- rep(rep(dimnames(FreqResultsStore_counting$site_years)[[2]], 
                              each = 10), 93)
years_short_format <- rep(rep(1:10, 4), 93)
Data_long_format <- data.frame(freq = Freq[!ind_na], 
                               snp_id = as.factor(rep(1:93, each = 10 * 4))[!ind_na],
                               site = sites_short_format[!ind_na],
                               year = years_short_format[!ind_na])

library(reshape2) # For switching between long and wide format using dcast
Data_wide_format <- dcast(Data_long_format, site + year ~ snp_id, value.var = 'freq')

# Fit pooled 
fit_pooled <- lm(freq ~ site + year, data = Data_long_format) 
summary(fit_pooled) # Not significant
anova(fit_pooled) # Not significant

# Fit individually
fit_individual <- lm(as.matrix(Data_wide_format[,3:95]) ~ Data_wide_format$site + Data_wide_format$year) 
summary_fit_indv <- summary(fit_individual)
p_value_store <- t(sapply(summary_fit_indv, function(x){x$coefficients[-1,4]}))

# Coefficient significance
apply(p_value_store, 2, function(x){sum(x < 0.05)}) # Before bonferroni correction
apply(p_value_store, 2, function(x){sum(x < 0.05/93)}) # After bonferroni correction
rm(p_value_store)

# Anova method (This features in the final draft of the manuscript)
anova_fit <- manova(as.matrix(Data_wide_format[,3:95]) ~ Data_wide_format$year + Data_wide_format$site)
X <- summary.aov(anova_fit)           # univariate ANOVA tables 
p_value_store <- t(sapply(X, function(x) {x[1:2,5]})) 
apply(p_value_store, 2, function(x){sum(x < 0.05)}) # Year comes first, then clinic
apply(p_value_store, 2, function(x){sum(x < 0.05/93)})
```


**************************Show in New WindowClear OutputExpand/Collapse Output

Call:
lm(formula = MOI_quants_site_year["50%", ] ~ X_years)

Residuals:
     Min       1Q   Median       3Q      Max 
-0.29284 -0.04597  0.01826  0.05110  0.33251 

Coefficients:
             Estimate Std. Error t value Pr(>|t|)    
(Intercept)  1.564767   0.053311  29.352  < 2e-16 ***
X_years     -0.038298   0.007648  -5.007 5.18e-05 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.1156 on 22 degrees of freedom
Multiple R-squared:  0.5326,	Adjusted R-squared:  0.5114 
F-statistic: 25.07 on 1 and 22 DF,  p-value: 5.176e-05


Call:
lm(formula = MOI_quants_site_year["50%", ] ~ X_years + X_sites)

Residuals:
      Min        1Q    Median        3Q       Max 
-0.216462 -0.046135 -0.000627  0.046743  0.238760 

Coefficients:
             Estimate Std. Error t value Pr(>|t|)    
(Intercept)  1.569013   0.085078  18.442 1.38e-13 ***
X_years     -0.039213   0.007143  -5.490 2.69e-05 ***
X_sitesMKT  -0.079711   0.069477  -1.147    0.265    
X_sitesMLA   0.090417   0.071699   1.261    0.223    
X_sitesWPA   0.023960   0.071199   0.337    0.740    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.09653 on 19 degrees of freedom
Multiple R-squared:  0.7186,	Adjusted R-squared:  0.6594 
F-statistic: 12.13 on 4 and 19 DF,  p-value: 4.586e-05

R Console


Call:
lm(formula = MOI_quants_site_year["50%", ] ~ X_years)

Residuals:
     Min       1Q   Median       3Q      Max 
-0.29284 -0.04597  0.01826  0.05110  0.33251 

Coefficients:
             Estimate Std. Error t value Pr(>|t|)    
(Intercept)  1.564767   0.053311  29.352  < 2e-16 ***
X_years     -0.038298   0.007648  -5.007 5.18e-05 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.1156 on 22 degrees of freedom
Multiple R-squared:  0.5326,	Adjusted R-squared:  0.5114 
F-statistic: 25.07 on 1 and 22 DF,  p-value: 5.176e-05


Call:
lm(formula = MOI_quants_site_year["50%", ] ~ X_years + X_sites)

Residuals:
      Min        1Q    Median        3Q       Max 
-0.216462 -0.046135 -0.000627  0.046743  0.238760 

Coefficients:
             Estimate Std. Error t value Pr(>|t|)    
(Intercept)  1.569013   0.085078  18.442 1.38e-13 ***
X_years     -0.039213   0.007143  -5.490 2.69e-05 ***
X_sitesMKT  -0.079711   0.069477  -1.147    0.265    
X_sitesMLA   0.090417   0.071699   1.261    0.223    
X_sitesWPA   0.023960   0.071199   0.337    0.740    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.09653 on 19 degrees of freedom
Multiple R-squared:  0.7186,	Adjusted R-squared:  0.6594 
F-statistic: 12.13 on 4 and 19 DF,  p-value: 4.586e-05

Show in New WindowClear OutputExpand/Collapse Output

Show in New WindowClear OutputExpand/Collapse Output

Call:
lm(formula = freq ~ site + year, data = Data_long_format)

Residuals:
     Min       1Q   Median       3Q      Max 
-0.44896 -0.09301  0.00154  0.09724  0.32743 

Coefficients:
              Estimate Std. Error t value Pr(>|t|)    
(Intercept)  6.466e-01  1.232e-02  52.475   <2e-16 ***
siteMKT      2.700e-03  1.006e-02   0.268    0.788    
siteMLA      2.224e-04  1.038e-02   0.021    0.983    
siteWPA     -8.707e-05  1.031e-02  -0.008    0.993    
year         9.458e-04  1.035e-03   0.914    0.361    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.1348 on 2227 degrees of freedom
Multiple R-squared:  0.0004532,	Adjusted R-squared:  -0.001342 
F-statistic: 0.2524 on 4 and 2227 DF,  p-value: 0.9083

Analysis of Variance Table

Response: freq
            Df Sum Sq   Mean Sq F value Pr(>F)
site         3  0.003 0.0010546  0.0580 0.9817
year         1  0.015 0.0151932  0.8358 0.3607
Residuals 2227 40.484 0.0181788               
     Estimate    Std. Error       t value      Pr(>|t|) 
-2.685351e-02  4.688386e-03 -5.727666e+00  1.608018e-05 
[1] "11"
R Console


Call:
lm(formula = freq ~ site + year, data = Data_long_format)

Residuals:
     Min       1Q   Median       3Q      Max 
-0.44896 -0.09301  0.00154  0.09724  0.32743 

Coefficients:
              Estimate Std. Error t value Pr(>|t|)    
(Intercept)  6.466e-01  1.232e-02  52.475   <2e-16 ***
siteMKT      2.700e-03  1.006e-02   0.268    0.788    
siteMLA      2.224e-04  1.038e-02   0.021    0.983    
siteWPA     -8.707e-05  1.031e-02  -0.008    0.993    
year         9.458e-04  1.035e-03   0.914    0.361    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.1348 on 2227 degrees of freedom
Multiple R-squared:  0.0004532,	Adjusted R-squared:  -0.001342 
F-statistic: 0.2524 on 4 and 2227 DF,  p-value: 0.9083

Analysis of Variance Table

Response: freq
            Df Sum Sq   Mean Sq F value Pr(>F)
site         3  0.003 0.0010546  0.0580 0.9817
year         1  0.015 0.0151932  0.8358 0.3607
Residuals 2227 40.484 0.0181788               
     Estimate    Std. Error       t value      Pr(>|t|) 
-2.685351e-02  4.688386e-03 -5.727666e+00  1.608018e-05 
[1] "11"*************************

# Frequencies by site (test differences at individual snp-level using the data partioned by year)

```{r, cach = FALSE, echo = FALSE} 
# Plot against site
plot(NULL,ylim = c(0.2,1), xlim = c(0.5,4.4), ylab = 'Allele frequency', 
     xlab = '', bty = 'n', xaxt = 'n')
for(i in 1:93){
  X0 <- rnorm(n = length(frequency_quants_site[i,'50%',]), sd = 0.1)
  points(y = frequency_quants_site[i,'50%',], x = 1:4 + X0, pch = 20, 
         col = 'darkblue')  
  segments(y0 = frequency_quants_site[i,'2.5%',], 
           y1 = frequency_quants_site[i,'97.5%',], 
           x0 = 1:4+X0, x1 = 1:4+X0, lwd = 0.5, 
           col = adjustcolor('darkblue', alpha.f = 0.75))
  points(y = FreqResultsStore_counting$site[,i], x = 1:4 + X0, pch = 4, 
         col = 'darkblue') 
}
axis(side = 1, cex.axis = 1, at = 1:4, line = -1, tick = FALSE, 
     labels = c('Mawker Thai', 'Maela','Mae Kon Ken','Wang Pha'), )
legend(x = 0.8, y = 0.36, bty = 'n', pch = c(20, 4), 
       col = 'darkblue', 
       legend = c('model-based estimate','frequency-based estimate')) 

# Inferred: No significant differnce between sites dropping snp-id 
kruskal.test(list(frequency_quants_site[,'50%','MKT'],
                  frequency_quants_site[,'50%','MLA'],
                  frequency_quants_site[,'50%','MKK'],
                  frequency_quants_site[,'50%','WPA']))

# Proportion: 
kruskal.test(list(FreqResultsStore_counting$site['MKT',],
                  FreqResultsStore_counting$site['MLA',],
                  FreqResultsStore_counting$site['MKK',],
                  FreqResultsStore_counting$site['WPA',]))
```

# Do proportion frequencies fall within CIs of inferred frequencies? 
```{r, cach = FALSE, echo = FALSE, include = TRUE} 
# Over sites 
within_CI_store <- array(dim = c(93, length(sites)), dimnames = list(NULL, sites))
for(site in sites){
  for(i in 1:93){
    prop <- FreqResultsStore_counting$site[ifelse(site == 'PLU', 'MKK', site), i]
    min_inf <- frequency_quants_site[i,'2.5%',site]
    max_inf <- frequency_quants_site[i,'97.5%',site]
    within_CI_store[i, site] <- min_inf <= prop & prop <= max_inf
  }  
}
mean(within_CI_store)
colSums(within_CI_store)/93 # Well over half fall within 95% CIs

# Over sites and years
within_CI_store <- frequency_quants_site_year[,1,]
for(i in 1:93){
  for(j in 1:24){
    X <- dimnames(frequency_quants_site_year)[[3]][j]
    site <- strsplit(X, split = ":")[[1]][1]
    year <- strsplit(X, split = ":")[[1]][2]
    prop <- FreqResultsStore_counting$site_year[year, site, i]
    min_inf <- frequency_quants_site_year[i,'2.5%',j]
    max_inf <- frequency_quants_site_year[i,'97.5%',j]
    within_CI_store[i, j] <- min_inf <= prop & prop <= max_inf
  }  
}
mean(within_CI_store) # Well over half fall within 95% CIs
```

