---
title: "Barcode data preparation"
header-includes:
- \usepackage{float}
output: pdf_document
bibliography: /Users/aimeet/Documents/BroadLaptop/Bibtex/library.bib
---

```{r setup, include=FALSE}
# Default chunk options
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=TRUE, cache.comments = FALSE, fig.pos = 'H', fig.width=7, fig.height=7)
library(knitr) # For kable
rm(list = ls())
```

This document summarizes the data stored in Data_store_barcode and how they were processed. Consensus sequences were previously called for isolates considered monoclonal (isolates with 1-5 heteroallelic genotyping calls - Fig \ref{fig: het mono v multi}), replacing any het calls with 'N' (Fig \ref{fig: base for het}), akin to a sequencing error [@Nkhoma2013]. Hereafter minor/major SNPs are encoded 0 and 1, respectively, where minority/majority was based on proportions discarding mixed and ignoring missing. 

```{r}
rm(list = ls())

# Source function to extract major freq
source('../../FunctionFiles/extract_major_freq.R')

# Load data from Nkhoma,S.C. et al. (2013) Population genetic correlates of declining transmission in a human pathogen. Mol. Ecol., 22, 273–285.
Data <- read.table(file ='../../RawData/TableS3_Nkhoma_et_al.txt', header = TRUE, skip = 0, sep = "\t")

# Inspect first two rows as column vectors size 10
t(head(Data[,1:10], n = 2))
dim(Data)

# Remove NA columns corresponding to failed SNPs 
Data <- Data[,-c(99,100)]
```


@Nkhoma2013 report

* 1731 infections overall and here we have: `r nrow(Data)`
* 629 from Wang Pha and here: `r sum(Data$Location.code == 'WPA')` from WPA
* 396 from Maela and here: `r sum(Data$Location.code == 'MLA')` from MLA
* 146 from Mae Kon Ken and here: `r sum(Data$Location.code == 'PLU')` from PLU
* 560 from Mawker-Thai and here: `r sum(Data$Location.code == 'MKT')` from MKT 

To conclude: MLA (Maela); WPA (Wang Pha); PLU (Mae Kon Ken); MKT (Mawker Thai)


```{r}
# Rename 'PLU' with 'MKK' (verfied from )
old_site_code <- levels(Data$Location.code)
new_site_code <- old_site_code
new_site_code[new_site_code == 'PLU'] <- 'MKK'
levels(Data$Location.code) <- new_site_code
Data[Data$Location.code == 'PLU', 'Location.code'] <- 'MKK'
```


```{r}
# Which samples have missing data? All at bottom
which(rowSums(Data == 'N') > 0) 

# Extract year and add
Year <- do.call(rbind, strsplit(as.character(Data[,2]), split = "-"))[,3] 
Data <- cbind(Year, Data) 
rownames(Data) <- Data$Sample.ID

# Separate barcodes from meta-data in first six columns
SNPData <- Data[,-(1:6)]
rownames(SNPData) <- Data$Sample.ID
MetaData <- Data[,1:6]
rownames(MetaData) <- MetaData$Sample.ID

# Number of viable SNPs
numSNPs <- ncol(SNPData)
numSNPs

# Check all 93 SNPs are polymorphic and biallelic - Yes!
t(sapply(apply(SNPData, 2, unique), FUN = function(x){x[x != 'N']}))
```


```{r}
#=======================================================================================
# Has consensus been attributed to samples declared monoclonal but with >1 het calls?
#=======================================================================================

# Samples with het genotyping outcomes:
hetconcensus <- apply(Data, 1, FUN = function(x){any(grepl('/', x))}) 

# Samples labeled with >1 het genotyping outcome detected
het1plus <- Data[,'No..of.heterozygous.SNPs'] > 0 

# Samples declared multiclonal
MOI1plus <- Data[,'Clonality'] == 'M' 

# Are there any declared monoclonal (!MOI1plus) with >1 het genotyping outcome
sum(!MOI1plus & het1plus) # yes

# Are there any declared monoclonal (!MOI1plus) with >1 het genotyping outcome and '/'?
sum(!MOI1plus & het1plus & hetconcensus) # no
```

```{r, include = TRUE, fig.cap=paste("\\label{fig: het mono v multi}Number of heteroallelic calls for isolates considered mono and multi-genotype."), fig.width=5, fig.height=7}
par(mfrow = c(2,1), family = "serif", las = 2, pty = 'm')
barplot(table(Data[!MOI1plus & het1plus,'No..of.heterozygous.SNPs']),
        ylab = 'No of samples', xlab = 'No. of declared het calls', main = 'Declared monoclonal')
barplot(table(Data[MOI1plus & het1plus,'No..of.heterozygous.SNPs']), 
        ylab = 'No of samples',
        xlab = 'No. of declared het calls', 
        main = 'Declared multiclonal',
        cex.names = 0.5, las = 2)
```

```{r, include = TRUE, fig.cap=paste("\\label{fig: base for het}Number of 'N' calls for isolates considered monoclonal but with 1 or more het calls versus all other isolates."), fig.width=5, fig.height = 5}
# Consensus 'missing' or a base (A, T, G, or C)? Hard to say, because het SNP not flagged. However, all samples declared monoclonal with >1 het call have $\geq$ 1 missing, while many alternative samples have no missing data 
par(mfrow = c(1,1), family = "serif", las = 2, pty = 'm')

monoclonal_but_het <- table(rowSums(SNPData[!MOI1plus & het1plus,] == 'N'))
alternative <- table(rowSums(SNPData[!(!MOI1plus & het1plus),] == 'N'))
max_missing_per_sample <- max(as.numeric(names(monoclonal_but_het)), as.numeric(names(alternative)))
store <- matrix(0,nrow = 2, ncol = 1+ max_missing_per_sample, 
                dimnames = list(c('monoclonal but with one or more het call', 'other'),
                                0:max_missing_per_sample))
store['monoclonal but with one or more het call', names(monoclonal_but_het)] <- monoclonal_but_het/sum(monoclonal_but_het)
store['other',names(alternative)] <- alternative/sum(alternative)
par(mfrow = c(1,1))
barplot(store, beside = TRUE, ylab = 'Frequency', xlab = 'No. missing calls per sample',
        legend.text = TRUE, args.legend = list(bty = 'n'))
```


```{r}
# Distribution of the declared monoclonal but with 1-5 het over years
monoclonal_but_het_logical <- !MOI1plus & het1plus # All at the bottom of the data set

# Years in relative terms
barplot(table(as.numeric(Year[monoclonal_but_het_logical])) / table(as.numeric(Year)),
        xlab = 'Year', ylab = 'Frequency of monoclonal with 1-5 het')
```


```{r}
# Distribution of the declared monoclonal but with 1-5 het over sites 
barplot(table(Data$Location.code[monoclonal_but_het_logical]) / table(Data$Location.code),
        xlab = 'Year', ylab = 'Frequency of "monoclonal" but 1-5 het')
```

```{r}
## Number of het calls over time
require(vioplot)
plot(NULL, bty = 'n', pch = 20, ylim =  c(0,100), ylab = 'No. of het calls', xlab = 'Years', las = 2, xlim = c(0.5,10.5))

for(year in unique(Year)){
  vioplot(Data[Year == year,'No..of.heterozygous.SNPs'], at = as.numeric(year), add = TRUE, col = 'gray')
  points(y = mean(Data[Year == year,'No..of.heterozygous.SNPs']), x = as.numeric(year), pch = 16)
}
```

```{r}
### Number of het calls over sites
plot(NULL, bty = 'n', pch = 20, ylim =  c(0,100), ylab = 'No. of het calls', xlab = 'Sites', las = 2, xlim = c(0.5,4.5), xaxt = 'n')
axis(side = 1, at = 1:4, labels = unique(Data$Location.code))
cols <- rainbow(4); names(cols) <- unique(Data$Location.code)

for(site in unique(Data$Location.code)){
  vioplot(Data[Data$Location.code == site,'No..of.heterozygous.SNPs'], at =    which(unique(Data$Location.code) == site), add = TRUE, col = 'gray')
  points(y = mean(Data[Data$Location.code == site,'No..of.heterozygous.SNPs']), x = which(unique(Data$Location.code) == site), pch = 16, col = cols[site])
}
```

```{r}
### Number of het calls over sites and years 
plot(NULL, bty = 'n', pch = 20, ylim =  c(0,100), ylab = 'No. of het calls', 
     xlab = 'Year', las = 2, xlim = c(0.5,10.5))
cols <- rainbow(4); names(cols) <- unique(Data$Location.code)

for(year in unique(Year)){
  for(site in unique(Data$Location.code)){
    Data_het <- Data[Data$Location.code == site & Data$Year == year,'No..of.heterozygous.SNPs']
    if(length(Data_het) > 10){
      vioplot(Data_het, at = as.numeric(year), add = TRUE, col = adjustcolor(cols[site], alpha.f = 0.5))
    }
  }
}
legend('topleft', legend = unique(Data$Location.code), col = cols, 
       pch = 16, bty = 'n', y.intersp = 1.2, horiz = TRUE, cex = 0.8)
```

```{r, cache = FALSE}
# Multiclonal logical vector
multiclonal <- MOI1plus

# Missing logical vector (includes monoclonal but <6 het)
missing <- apply(Data, 1, FUN = function(x){any(x[-(1:6)] == 'N')})
```

Of the data imported from TableS3_Nkhoma_et_al.txt

* Percentage declared multiclonal: `r round(100*(sum(multiclonal)/length(multiclonal)),2)`
* Percentage with missing: `r round(100*(sum(missing)/length(missing)),2)`
* Number with missing: `r sum(missing)`
* Number with missing excluding the samples with $\leq 5$ het calls: 
`r sum(missing) - sum(!MOI1plus & het1plus)`


```{r, cache = FALSE}
# Filter het and missing
Data_no_multiclonal_missing <- Data[!missing & !multiclonal,]
Data_no_multiclonal <- Data[!multiclonal,]
Data_no_missing <- Data[!missing,]

# Extract major based on non-multiclonal only frequencies
majorSNPstore <- do.call(rbind, apply(Data_no_multiclonal[,-(1:6)], 2, extract_major_freq))
```


```{r}
#==============================================================================================
# Create matrices with major (1) and minor (0), and freq matrix 
#==============================================================================================

# Create Freq stores
SNPDataFreq_no_missing <- SNPData_no_missing <- Data_no_missing[,-(1:6)];
SNPDataFreq_no_multiclonal <- SNPData_no_multiclonal <- Data_no_multiclonal[,-(1:6)];
SNPDataFreq_no_multiclonal_missing <- SNPData_no_multiclonal_missing <- Data_no_multiclonal_missing[,-(1:6)]

# Populate stores with minor and NA frequencies
SNPData[] <- 0 # All minor (0)
SNPDataFreq_no_missing[] <- SNPData_no_missing[] <- NA
SNPDataFreq_no_multiclonal[] <- SNPData_no_multiclonal[] <- NA
SNPDataFreq_no_multiclonal_missing[] <- SNPData_no_multiclonal_missing[] <- NA

for(j in 1:numSNPs){
  
  # Populate matrix inc mixed and missing (replace missing outside loop)
  indx_major <- (Data[,-(1:6)][,j] == majorSNPstore[j][[1]])
  indx_mixed <- grepl('/',Data[,-(1:6)][,j])
  SNPData[indx_major,j] <- 1 # Major 
  SNPData[indx_mixed,j] <- 0.5 # Mixed
  
  # Populate matrices with no mixed or missing 
  indx_no_missing <- (Data_no_missing[,-(1:6)][,j] == majorSNPstore[j][[1]])
  indx_no_multiclonal <- (Data_no_multiclonal[,-(1:6)][,j] == majorSNPstore[j][[1]])
  indx_no_multiclonal_missing <- (Data_no_multiclonal_missing[,-(1:6)][,j] == majorSNPstore[j][[1]])
  
  # Populate matrices with no mixed or missing
  SNPDataFreq_no_multiclonal_missing[indx_no_multiclonal_missing,j] <- majorSNPstore[j,][[2]]
  SNPDataFreq_no_multiclonal_missing[!indx_no_multiclonal_missing,j] <-  (1 - majorSNPstore[j,][[2]])
  SNPData_no_multiclonal_missing[indx_no_multiclonal_missing,j] <- 1 # Major
  SNPData_no_multiclonal_missing[!indx_no_multiclonal_missing,j] <- 0 # Minor
  
  # Populate matrices with no mixed (replace missing outside loop)
  SNPDataFreq_no_multiclonal[indx_no_multiclonal,j] <-  majorSNPstore[j,][[2]]
  SNPDataFreq_no_multiclonal[!indx_no_multiclonal,j] <-  (1 - majorSNPstore[j,][[2]])
  SNPData_no_multiclonal[indx_no_multiclonal,j] <- 1
  SNPData_no_multiclonal[!indx_no_multiclonal,j] <- 0
  
  # Populate matrices with no missing (replace multiclonal outside loop) 
  SNPDataFreq_no_missing[indx_no_multiclonal,j] <-  majorSNPstore[j,][[2]]
  SNPDataFreq_no_missing[!indx_no_multiclonal,j] <-  (1 - majorSNPstore[j,][[2]])
  SNPData_no_missing[indx_no_multiclonal,j] <- 1
  SNPData_no_missing[!indx_no_multiclonal,j] <- 0
  # Add mixed
  indx_mixed_no_missing <- grepl('/',Data_no_missing[,-(1:6)][,j])
  SNPData_no_missing[indx_mixed_no_missing,j] <- 0.5
}

# Reintroduce missing for SNPData and SNPData_no_multiclonal
SNPData[Data[,-(1:6)] == 'N'] <- NA
SNPDataFreq_no_multiclonal[Data_no_multiclonal[,-(1:6)] == 'N'] <- NA
SNPData_no_multiclonal[Data_no_multiclonal[,-(1:6)] == 'N'] <- NA
```

```{r}
# All data (all missing at the top = bottom of data set):
# * Missing = black
# * Minor = dark grey
# * Mixed = light grey 
# * Major = white

# Binary SNP data....
SNPData_temp <- SNPData
SNPData_temp[is.na(SNPData_temp)] <- -1
image(z = t(SNPData_temp), xaxt = 'n', yaxt = 'n', zlim = c(-1,1),
           col = grey.colors(4, start = 0, end = 1))
```

```{r, cache = FALSE, echo = FALSE}
# No mixed (all missing at the top = bottom of data set):
# 
# * Missing = black
# * Minor = grey
# * Major = white

SNPData_no_multiclonal_temp <- SNPData_no_multiclonal
SNPData_no_multiclonal_temp[is.na(SNPData_no_multiclonal_temp)] <- -1
image(z = t(SNPData_no_multiclonal_temp), xaxt = 'n', yaxt = 'n', zlim = c(-1,1),
           col = grey.colors(3, start = 0, end = 1))
```

```{r, cache = FALSE, echo = FALSE}
# No missing
image(z = t(SNPData_no_multiclonal_missing), xaxt = 'n', yaxt = 'n', zlim = c(-1,1),
           col = grey.colors(3, start = 0, end = 1))
```

```{r}
par(mfrow = c(1,1))
# Freq estimates for no multiclonal data
SNPDataFreq_no_multiclonal_temp <- SNPDataFreq_no_multiclonal
SNPDataFreq_no_multiclonal_temp[is.na(SNPDataFreq_no_multiclonal_temp)] <- -0.1
image(z = t(SNPDataFreq_no_multiclonal_temp), xaxt = 'n', yaxt = 'n', zlim = c(-0.1,1),
           col = grey.colors(100, start = 0, end = 1))
```

```{r}
# Freq estimates no missing
image(z = t(SNPDataFreq_no_multiclonal_missing), xaxt = 'n', yaxt = 'n', zlim = c(-0.1,1),
           col = grey.colors(100, start = 0, end = 1))
```


```{r, cache = FALSE}
# =========================================================================
# Save data matrices
# =========================================================================
Data_store <- list(Data = Data,
                        Data_no_multiclonal_missing = Data_no_multiclonal_missing,
                        Data_no_multiclonal = Data_no_multiclonal,
                        Data_no_missing = Data_no_missing,
                        MetaData = MetaData,
                        SNPData = SNPData,
                        SNPData_no_missing = SNPData_no_missing,
                        SNPData_no_multiclonal = SNPData_no_multiclonal,
                        SNPData_no_multiclonal_missing = SNPData_no_multiclonal_missing,
                        SNPDataFreq_no_missing = SNPDataFreq_no_missing,
                        SNPDataFreq_no_multiclonal = SNPDataFreq_no_multiclonal,
                        SNPDataFreq_no_multiclonal_missing = SNPDataFreq_no_multiclonal_missing)
save(Data_store, file = '../../RData/Data_store_Barcode.RData')

# Data for hmmIBD, which inc. single-genotype only, are generated and saved in S4 Appendix
```

In the analyses reported in the main manuscript, which are all based on either Data_store_barcode.RData, Barcode93.txt or Barcode24.txt (generated in S4Appendix.Rmd), we use single-genotype isolates only. 

## Sample sizes
```{r, cache = FALSE}
generate_sample_size_matrix <- function(Data){
  sample_size_store <- matrix(NA, 
                              nrow = length(unique(Data$Location.code)), 
                              ncol = length(unique(Data$Year)), 
                              dimnames = list(unique(Data$Location.code), 
                                              unique(Data$Year)))
  
  for(site in unique(Data$Location.code)){
    for(year in unique(Data$Year)){
      sample_size_store[site, year] <- sum(Data$Year == year & Data$Location.code == site)
    }
  }
  
  # And row and column sums
  X0 <- cbind(sample_size_store, Total = rowSums(sample_size_store))
  X1 <- rbind(X0, Total = colSums(X0))
  
  return(X1)
}

X1 <- generate_sample_size_matrix(Data = Data_no_missing)
X2 <- generate_sample_size_matrix(Data = Data_no_multiclonal)
X3 <- generate_sample_size_matrix(Data = Data_no_multiclonal_missing)
X4 <- generate_sample_size_matrix(Data = Data)

# Save X2 since that is the one we are ultimately working with
Sample_size_table = X2
save(Sample_size_table, file = '../../RData/Sample_size_table_Barcode.RData')
```

```{r, include = TRUE} 
kable(X1, caption = '\\label{tab:x1}Isolate counts over sites and years, excluding those with missing data')
```

```{r, include = TRUE} 
kable(X2, caption = '\\label{tab:x2}Single-genotype isolate counts over sites and years')
```

```{r, include = TRUE} 
kable(X3, caption = '\\label{tab:x3}Single-genotype isolate counts over sites and years, excluding those with missing data')
```

```{r, include = TRUE} 
kable(X4, caption = '\\label{tab:x4}Multi and single-genotype isolate counts over sites and years')
```